`gosset`: An R package for analysis and synthesis of ranking data in agricultural experimentation

### 2022-06-22 inst/paper/paper.Rmd
updating

---
title: '`gosset`: An R package for analysis and synthesis of ranking data in agricultural experimentation'
abstract: null
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: 72
bibliography: paper.bib
header-includes: \usepackage{caption} \captionsetup[figure]{labelformat=empty}
---

@@

: *Code metadata (mandatory)*

@@

* `plot` provides an alternative to the default plot function of the PlackettLuce package.

* `regret` computes the regret [@Loomes_1982;  ].

* `reliability` computes the reliability of crop varieties, the probability to outperform the reference variety [@eskridge_1992].

* `worth_bar`  creates a bar plot of the estimated worth values for each evaluated variety.

* `worth_map` creates a heatmap plot of the estimated worth values for all varieties considering each of the
evaluated traits.

# Illustrative Examples

To demonstrate the functionality of the gosset package, we use the `nicabean` dataset, which was generated with decentralized on-farm trials of common bean (*Phaseolus vulgaris* L.) varieties in Nicaragua over five seasons (between 2015 and 2016). Following the tricot  approach [@vanetten_beza_2019], farmers were asked to test in their farms three varieties of common bean. The varieties were randomly assigned as incomplete blocks of size three (out of 10 varieties). The farmers assessed which of the three varieties has the best and worst performance in nine traits (vigor, architecture, resistance to pests, resistance to diseases, tolerance to drought, yield, marketability, taste) and overall appreciation). The farmers also provided their overall appreciation about the varieties, i.e., which variety has the best and the worst performance based on the overall performance considering all the traits. We used the Plackett-Luce model implemented in the R package *PlackettLuce* [@Turner2020].

First, we load the required packages and data. The `nicabean` dataset contains the tricot rankings (`trial`), the climate covariates (`covar`) and the traits observations (`traits`).

@@

Make a PlackettLuce rank using the function `rank_numeric`.

@@

The following lines show the first 10 rankings for the trait vigor, 
in the format used by the PlackettLuce package.

```{r fetch, message = FALSE, eval = TRUE, echo = TRUE}
head(R[[1]], 5)
```

The correlation among rankings can be assessed using the function `kendallTau`.

```{r}
baseline <- which(grepl("OverallAppreciation", traits))

@@

kendall$trait <- traits[-baseline]

print(kendall)
```

@@

```{r, message=FALSE, eval=TRUE, echo=TRUE}
worth_map(mod[-baseline],
          labels = traits[-baseline], 
          ref = "Amadeus 77") +
  labs(x = "Variety",
       y = "Trait")
```

To consider the effect of climate factors on variety performance, we used the climatic covariates to fit a 
Plackett-Luce tree. For brevity, we used the variables maximum nighttime temperature (maxNT) and  maximum daytime temperature, but more variables can be used in a Plackett-Luce tree. Temperature data were obtained from the NASA POWER dataset, using the R package *nasapower* [@nasapower]. The temperature data downloaded from NASA POWER were used for
computing the climatic variables with the package *climatrends* [@climatrendspkg]. In this case, we only use the rankings
corresponding to the overall performance of varieties.

@@

**Zenodo repository for the gosset package:** <https://doi.org/10.5281/zenodo.6339989>

::: thebibliography
:::

*Please add the reference to the software repository if DOI for software is available.*

>>>>

---
title: '`gosset`: An R package for analysis and synthesis of ranking data in agricultural experimentation'
abstract: null
output:
   pdf_document:
     fig_caption: true
     number_sections: false
     keep_tex: false
latex_engine: pdflatex
#output:
# word_document: default
editor_options:
  markdown:
    wrap: 72
bibliography: paper.bib
csl: citationstyle.csl
header-includes:
  - \usepackage{pdfpages}
  - \usepackage{caption}
  - \usepackage{xcolor}
  - \captionsetup[figure]{font=small}
  - \captionsetup[table]{font=small}
---

```{r setup_opts, include=FALSE}
library("knitr")
knitr::opts_chunk$set(echo = FALSE,
                      error = FALSE,
                      message=FALSE,
                      warning = FALSE)
```

@@

: *Code metadata*

@@

* `plot` provides an alternative method to the default plot function of the PlackettLuce package.  
* `regret` computes the regret [@Loomes_1982;  ].  
* `reliability` computes the reliability of crop varieties, the probability to outperform the reference variety [@eskridge_1992].  
* `worth_bar`  creates a bar plot of the estimated worth values for each evaluated variety.  
* `worth_map` creates a heatmap plot of the estimated worth values for all varieties considering each of the
evaluated traits.

# Illustrative example

To demonstrate the functionality of the gosset package, we use the `nicabean` dataset, which was generated with decentralized on-farm trials of common bean (*Phaseolus vulgaris* L.) varieties in Nicaragua over five seasons (between 2015 and 2016). Following the tricot  approach [@vanetten_beza_2019], farmers were asked to test in their farms three varieties of common bean. The varieties were randomly assigned as incomplete blocks of size three (out of 10 varieties). The farmers assessed which of the three varieties had the best and worst performance in eight traits (vigor, architecture, resistance to pests, resistance to diseases, tolerance to drought, yield, marketability, taste). The farmers also provided their overall appreciation about the varieties, i.e., which variety had the best and the worst performance based on the overall performance considering all the traits. To analyze the data, we use the Plackett-Luce model implemented in the R package *PlackettLuce* [@Turner2020].

The `nicabean` is a list with two data frames: i) `trial`, contains the trial data with farmers' evaluations, ranked from 1 to 3, with 1 being the higher ranked variety and 3 the lowest ranked variety for the given trait and incomplete block; ii) `covar`, contains the covariates associated to the on-farm trial plots and farmers. This example will require the packages PlackettLuce [@Turner2020], climatrends [@climatrends], nasapower [@nasapower] and ggplot2 [@ggplot2].

@@

To start the analysis of the data, we transform the rankings into Plackett-Luce rankings using the function `rank_numeric()`. We run iteratively over the traits adding the rankings to a list called `R`. Since the varieties are ranked in an ascending order, with 1 being the higher ranked and 3 the lower ranked, we use the argument `asceding = TRUE` to indicate which order should be used.

@@

Then, using the function `kendallTau()` we can assess the correlation between rankings. This approach can be support at least two decision making processes by, for example i) indicating the traits

```{r kendall1, message=FALSE, eval=TRUE, echo=TRUE}
baseline <- which(grepl("OverallAppreciation", traits))

@@

kendall$trait <- traits[-baseline]
```

```{r kendall2, message=FALSE, eval=TRUE, echo=FALSE}

kendall <- kendall[,c(3, 1)]
kendall[,2] <- round(kendall[,2], 3)

kable(kendall, 
          caption = "Kendall correlation between 'overall performance' and the other traits assessed in the Nicaragua bean on-farm trials.",
          align = "l",
          row.names = FALSE)

@@

```{r, message=FALSE, eval=TRUE, echo=TRUE, fig.cap="Trait performance (log-worth) of bean varieties in Niragua. Variety 'Amadeus' is set as reference (log-worth = 0). Blue values indicate a superior performance of varieties for a given trait, compared to the reference.  Red values indicate a variety with weak performance for the given trait, compared to the reference."}
worth_map(mod[-baseline],
          labels = traits[-baseline], 
          ref = "Amadeus 77") +
  labs(x = "Variety",
       y = "Trait")
```

To consider the effect of climate factors on variety performance, we used the climatic covariates to fit a 
Plackett-Luce tree. For brevity, we used the variables maximum nighttime temperature (maxNT) and  maximum daytime temperature, but more variables can be used in a Plackett-Luce tree. Temperature data were obtained from the NASA POWER dataset, using the R package *nasapower* [@nasapower]. The temperature data downloaded from NASA POWER were used for computing the climatic variables with the package *climatrends* [@climatrends]. In this case, we only use the rankings corresponding to the overall performance of varieties.

@@

::: {#refs}
:::
